Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-21 main.tex
rule permutation

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$ other than $\aiD$ and $\forall$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\juihsuan{For this, we need the introduction rule of $\fff$ in $\FOMLL$
(consider the $\aiD$ rule). But if we add it, we'll need the
$\mathsf{ai}\downarrow$ rule to have Lemma \ref{lem:MLL1->MLS1}}
\lutz{shit. so, we need another way to prove lemma~\ref{lem:MLS1->MLL1}, I'd like to avoid $\fff$-intro.}
\juihsuan{This lemma holds for all the rules except $\aiD$ (we may need to
mention that $A \fequ B \Rightarrow \ \vdash_{\FOMLL} \dual{B}, A$). The inductive
step for $\aiD$ in the proof of Lemma \ref{lem:MLS1->MLL1} can be solved by
replacing $t$ with $a \vlor \dual{a}$ and the $\ttt$-rule with the $\ax$
rule followed by the $\vlor$ rule}\lutz{yes! and similar for $\forall$. Can you do that?}

\lutz{actually NO! we can prove $\sqn{\fff,\ttt}$ using $\ax$ without a special rule for $\fff$. So can you please undo the changes and delete the comments?}
\begin{proof}
  This is a straightforward exercise that we leave to the reader. (Note that the $\ax$-rule can be applied to $\sqn{\fff,\ttt}$ in the cases of $\rr \in \set{\aiD, \faD}$.)
\end{proof}

@@

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof $\Proof$ of $\sqn{\Scons A}$.
  We now make a case analysis on $\rr$:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}$ :
      In this case, we replace simultaneously in $\Pi$ the corresponding
      occurrences of $\ttt$ with $a \vlor \cneg a$ and the rule instance
      $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
      $
        \vlderivation{
          \vlin{\vlor}{}{\sqn{a \vlor \cneg a}}{
            \vlin{\axr}{}{\sqn{a, \cneg a}}{
              \vlhy{}}}}
      $, and we get a proof $\Pi'$ of $\sqn{\Scons{a \vlor \cneg a}}$.
    \item $\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$
      In this case, we replace simultaneously in $\Pi$ the corresponding
      occurrences of $\ttt$ with $\forall x.\ttt$ and the rule instance
      $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
      $
      \vlderivation{
        \vlin{\forall}{}{\sqn{\forall x.\ttt}}{
          \vlin{\axr}{}{\sqn{\ttt}}{
            \vlhy{}}}}
      $, and we get a proof $\Pi'$ of $\sqn{\Scons{\forall x.\ttt}}$.
    \item $\rr \notin \set{\aiD, \faD}$ :
      By Lemma~\ref{lem:context}, we have a $\FOMLL$ proof of
      $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
      \begin{equation*}
        \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
      \end{equation*}
      and then apply Theorem~\ref{thm:cutelim}.
  \end{itemize}
  \tocheck{}
  \end{proof}

@@

First, if an instance $\vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}$
 of $\weakr$ is followed by a rule in which $A$ is not in the principal
formula, it can be permuted downwards.
Otherwise, the proof can be transformed using the following rewriting rules.

@@

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}

>>>>

\begin{lemma}\label{lem:shallow}
  Let $\vlinf{\rr}{}{\Scons B}{\Scons A}$ be an inference rule in
  $\FOMLS$ other than $\aiD$. Then the sequent $\sqn{\cneg A,B}$ is provable in $\FOMLL$.
\end{lemma}
\begin{proof}
  This is a straightforward exercise that we leave to the reader. (Note that the
$\ax$-rule can be applied to $\sqn{\fff,\ttt}$ in the cases of $\rr = \faD$.)
\end{proof}

@@

\begin{proof}
  We proceed by induction on the number of inference steps in the
  proof of $C$ in $\FOMLS$. Consider the bottommost rule instance
  $\vlinf{\rr}{}{\Scons B}{\Scons A}$. By induction hypothesis we have
  a $\FOMLL$ proof $\Proof$ of $\sqn{\Scons A}$.
  We now make a case analysis on $\rr$:
  \begin{itemize}
    \setlength\itemsep{.5em}
    \item $\vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}$ :
      In this case, we replace simultaneously in $\Pi$ the corresponding
      occurrences of $\ttt$ with $a \vlor \cneg a$ and the rule instance
      $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
      $
        \vlderivation{
          \vlin{\vlor}{}{\sqn{a \vlor \cneg a}}{
            \vlin{\axr}{}{\sqn{a, \cneg a}}{
              \vlhy{}}}}
      $, and we get a proof $\Pi'$ of $\sqn{\Scons{a \vlor \cneg a}}$.
    %\item $\vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}$
     % In this case, we replace simultaneously in $\Pi$ the corresponding
     % occurrences of $\ttt$ with $\forall x.\ttt$ and the rule instance
     % $\vlinf{\ttt}{}{\sqn{\ttt}}{}$ with the derivation
     % $
     % \vlderivation{
     %   \vlin{\forall}{}{\sqn{\forall x.\ttt}}{
     %     \vlin{\axr}{}{\sqn{\ttt}}{
     %       \vlhy{}}}}
     % $, and we get a proof $\Pi'$ of $\sqn{\Scons{\forall x.\ttt}}$.
    \item In the other cases,
      by Lemma~\ref{lem:context}, we have a $\FOMLL$ proof of
      $\sqn{\cneg{\Scons A},\Scons B}$. We can compose them via $\cutr$:
      \begin{equation*}
        \vliinf{\cutr}{}{\sqn{\Scons B}}{\sqn{\Scons A}}{\sqn{\cneg{\Scons
          A},\Scons B}}
      \end{equation*}
      and then apply Theorem~\ref{thm:cutelim}.
  \end{itemize}
  \tocheck{}
  \end{proof}

@@

First, if an instance of $\vlinf{\weakr}{}{\sqn{\Gamma,A}}{\sqn{\Gamma}}$
 is followed by a rule in which $A$ is not in the principal
formula, it can be permuted downwards.
Otherwise, the proof can be transformed using the following rewriting rules.

@@

\begin{proof}
  First, observe that the $\exists$ rule can be permuted downwards over all the
other rules since $A\ssubst{x}{t}$ has the same structure as $A$ and none of the
other rules has a premise of the form $\Scons{\exists x.A}$. It suffices now to
prove that for all $r_1 \in \set{\forall, \aiD, \tttD}$, for all $r_2 \in
\set{\sw, \mix, \fequ}$, we can permute $r_1$ upwards over $r_2$.
  We show some cases here, and leave the other to the reader.
  \begin{equation*}
  \vlderivation{
    \vlin{\aiD}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\sw}{}{\Scons{(A \vlan \ttt) \vlor C}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \leadsto
  \vlderivation{
    \vlin{\sw}{}{\Scons{(A \vlan (a \vlor \cneg a)) \vlor C}}{
      \vlin{\aiD}{}{\Scons{A \vlan ((a \vlor \cneg a) \vlor C)}}{
        \vlhy{\Scons{A \vlan (\ttt \vlor C)}}}}}
  \end{equation*}
  \begin{equation*}
    \vlderivation{
      \vlin{\tttD}{}{\Scons{(A \vlor (B \vlan \ttt))}}{
        \vlin{\mix}{}{\Scons{A \vlor B}}{
          \vlhy{\Scons{A \vlan B}}}}}
  \leadsto
  \vlderivation{
    \vlin{\mix}{}{\Scons{(A \vlor (B \vlan \ttt)}}{
      \vlin{\tttD}{}{\Scons{A \vlan (B \vlan \ttt)}}{
        \vlhy{\Scons{A \vlan B}}}}}
  \end{equation*}

  \tocheck{}
\end{proof}

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wrD,\acD,\cfaD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{B}{
        \vlde{\set{\acD,\cfaD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}
